---
title: "Reproducible Research Project 1"
author: "Faraz Shaikh"
date: "7/5/2020"
output: html_document
---

```{r}
getwd()
knitr::opts_chunk$set(root.dir = normalizePath("E:/R"))
getwd()
knitr::opts_chunk$get("root.dir")
```

In this project, we've instructions:

i. Code for reading in the dataset and/or processing the data
ii. Histogram of the total number of steps taken each day
iii. Mean and median number of steps taken each day
iv. Time series plot of the average number of steps taken
v. The 5-minute interval that, on average, contains the maximum number of steps
vi. Code to describe and show a strategy for imputing missing data
vii. Histogram of the total number of steps taken each day after missing values are imputed
viii. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
ix. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

## **i. Code for reading in the dataset and/or processing the data**
```{r, echo = TRUE}
setwd("E:/R")
project <- read.csv("activity.csv")
```

Exploring the dataset
```{r}
head(project)
tail(project)
names(project)
str(project)
dim(project)
# Finding total missing data
sum(is.na(project$steps))/dim(project)[[1]]
#Transformation to date format from data column using Lubridate
library(lubridate)
project$date <- ymd(project$date)
length(unique(project$date))
```

## **ii. Histogram of the total number of steps taken each day**
```{r, echo = TRUE}
library(ggplot2)
Q2 <- data.frame(tapply(project$steps, project$date, sum, na.rm = TRUE))
Q2$date <- rownames(Q2)
rownames(Q2) <- NULL
names(Q2)[[1]] <- "Total Steps"
png("assignment_plot1.png")
# Bar chart for Total steps by date 
ggplot(Q2,aes(y=`Total Steps`,x=`date`))+geom_bar(stat="identity") + ylab("Total Steps")+xlab("Date")+ggtitle("Total Steps by date")
dev.off()
ggplot(Q2,aes(y=`Total Steps`,x=`date`))+geom_bar(stat="identity") + ylab("Total Steps")+xlab("Date")+ggtitle("Total Steps by date")
# Total Steps in Histogram
qplot(Q2$`Total Steps`,geom="histogram",xlab="Total Steps",ylab="Counts",main="Total Steps Histogram")
png("assignment_plot1.1.png")
qplot(Q2$`Total Steps`,geom="histogram",xlab="Total Steps",ylab="Counts",main="Total Steps Histogram")
dev.off()
```

## **iii. Mean and median number of steps taken each day**
```{r, echo=TRUE}
library(dplyr)
Q3 <- data.frame(round(tapply(project$steps, project$date, sum, na.rm = TRUE),2))
Q3$date <- rownames(Q3)
rownames(Q3) <- NULL
names(Q3)[[1]] <- "Mean Steps"
temp <- project%>%select(date,steps) %>% group_by(date) %>% summarize(median(steps))
names(temp)[[2]] <- "Median Steps"
Q3$median <- temp$`Median Steps`
Q3 <- Q3 %>% select(date, `Mean Steps`, median)

```

## **iv. Time series plot of the average number of steps taken**
```{r,echo=TRUE}
Q4 <- Q3
Q4$date <- as.Date(Q4$date, format = "%Y-%M-%d")
ggplot(Q4,aes(x=date,y=`Mean Steps`))+geom_bar(stat="identity")+scale_x_date()+ylab("Mean Steps daily")+xlab("Date")+ggtitle("Mean Steps By date")
png("Assignment_plot4.png")
ggplot(Q4,aes(x=date,y=`Mean Steps`))+geom_bar(stat="identity")+scale_x_date()+ylab("Mean Steps daily")+xlab("Date")+ggtitle("Mean Steps By date")
dev.off()

```

## **v. The 5-minute interval that, on average, contains the maximum number of steps**
```{r, echo=TRUE}
project$interval <- factor(project$interval)
Q5 <- aggregate(data = project, steps~date+interval, FUN = "mean")
Q5 <- aggregate(data = Q5, steps~interval, FUN = "max")
```

## **vi. Code to describe and show a strategy for imputing missing data**
```{r, echo = TRUE}
Q6 <- project
Q6$missing <- is.na(Q6$steps)
Q6<-aggregate(data=Q6,missing~date+interval,FUN="sum")
Q6.1 <- data.frame(tapply(Q6$missing, Q6$date, sum))
Q6.1$date <- rownames(Q6.1)
rownames(Q6.1) <- NULL
names(Q6.1) <- c("missing", "date")
Q6.1$date <- as.Date(Q6.1$date, format = "%Y-%M-%d")

Q6.2 <- data.frame(tapply(Q6$missing, Q6$interval, sum))
Q6.2$date <- rownames(Q6.2)
rownames(Q6.2) <- NULL
names(Q6.2) <- c("missing", "Interval")

#par(mfrow = c(1,2))
plot(y=Q6.1$missing,x=Q6.1$date,main="Missing values disributed by Date")
#plot(y=Q6.1$missing,x=Q6.1$Interval,main="Missing values distributed by Interval")
table(project$date)

```

```{r, echo=TRUE}
library(lubridate)
Q6.3 <- as.data.frame(Q6.1) %>% select(date, missing) %>% arrange(desc(missing))
Q6.3 <- Q6.3[which(Q6.3$missing!=0),]
Q6.3$weekdays <- wday(Q6.3$date, label = TRUE)

Q6.4 <- project
Q6.4$weekdays  <- wday(Q6.4$date, label = TRUE)
#The mean of steps on every monday and every interval.
Q6.5<- aggregate(data = Q6.4, steps~interval+weekdays,FUN="mean",na.rm=TRUE)
# Merge the pre-imputation table Q6.4 with average table Q6.5
Q6.6 <- merge(x=Q6.4, y=Q6.5,by.x=c("interval","weekdays"),by.y=c("interval","weekdays"),all.x = TRUE)

Q6.6$Steps.Updated <- 0
for (i in 1:dim(Q6.6)[[1]]) {
if(is.na(Q6.6[i,3])){Q6.6[i,6]=Q6.6[i,5]}
else {Q6.6[i,6]=Q6.6[i,3]}  

}
#Simplifying
Q6.6 <- Q6.6 %>% select(date, weekdays, interval, Steps.Updated)
names(Q6.6)[[4]] <- "Steps"


```

## **vii. Histogram of the total number of steps taken each day after missing values are imputed**
```{r, echo = TRUE}
png("assignment_plot7.png")
qplot(Q6.6$Steps,geom="histogram",main="Total steps taken after fixing missing values",xlab="Steps",ylab="Count")
dev.off()
qplot(Q6.6$Steps,geom="histogram",main="Total steps taken after fixing missing values",xlab="Steps",ylab="Count")

```

## **viii. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends**
```{r, echo = TRUE}
Q8 <- Q6.6
levels(Q8$weekdays) <- c(1,2,3,4,5,6,7)
Q8$WDWE <- Q8$weekdays %in% c(1,2,3,4,5)
Q8.1 <- aggregate(date=Q8, steps~interval+WDWE, mean, na.rm = TRUE)
Q8.1$WDWE <- as.factor(Q8.1$WDWE)
levels(Q8.1$WDWE) <- c("Weekend", "Weekday")
png("assignment_plot8.png")
ggplot(data=Q8.1,aes(y=Steps,x=interval,group=1,color=WDWE))+geom_line()+scale_x_discrete(breaks=seq(0,250,by=300))+ylab("Mean Steps")+xlab("Interval")+ggtitle("Mean Steps distribution by Interval")
dev.off()
ggplot(data=Q8.1,aes(y=Steps,x=interval,group=1,color=WDWE))+geom_line()+scale_x_discrete(breaks=seq(0,250,by=300))+ylab("Mean Steps")+xlab("Interval")+ggtitle("Mean Steps distribution by Interval")

# Panel plot
Q8.1$interval <- as.numeric(as.character(Q8.1$interval))
library(lattice)
xyplot(data=Q8.1,Steps~interval|WDWE,grid=TRUE,type=c("p", "smooth"),lwd=4,panel=panel.smoothScatter)
library(hexbin)
hexbinplot(data=Q8.1,Steps~interval|WDWE,aspect=1,bins=50)
png("assignment_plot8.1.png")
xyplot(data=Q8.1,Steps~interval|WDWE,grid=TRUE,type=c("p", "smooth"),lwd=4,panel=panel.smoothScatter)
dev.off()

png("assignment_plot8.2.png")
hexbinplot(data = Q8.1, Steps~interval|WDWE, aspect=1, bins = 50)
dev.off()
```

## **The End**